---
title       : Coursera - Developing Data Products
subtitle    : Programming Assignment to Create Shiny Application
author      : David Mullins
job         : 
framework   : html5slides        # {io2012, html5slides, shower, dzslides, ...}
highlighter : prettify
hitheme     : tomorrow      # 
widgets     : []            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

# Coursera - Developing Data Products

## Programming Assignment to Create Shiny Application

### *David Mullins*

#### 01/08/2019

--- .class #id 

## Shiny app based on the 'Titanic' sample dataset available in base R

This is a simple dataset with 4 'predictor' categorical variables and 1 'response' variable in the frequency of survival.

```{r dataset_head}
summary(as.data.frame(Titanic))
```

---

## Choose travel class

Radio buttons were decided on as a means of selecting between different travel classes. The shiny code for this is shown.

```{r radio_buttons, eval=FALSE}
radioButtons(
    inputId = "Class",
    label = "Pick the class of travel:",
    choices = list(
        `1st` = "1st",
        `2nd` = "2nd",
        `3rd` = "3rd",
        `Crew` = "Crew"
    )
)
```

---

## The plot was then created with ggplot2

```{r ggplot2, eval=FALSE}
outPlot <-
    ggplot(data = plot_data.df, aes(x = xaxis, y = chances, label = paste(round(chances*100,1), "%", sep=" ")) ) +
        ## Area plot which will be filled under the line
        geom_area(fill = "skyblue", colour = "skyblue") +
        ## Set up y-axis
        geom_hline( aes(yintercept = 1), colour = "black" ) +
        scale_y_continuous(name = "variable", limits = c(0, 1)) +
        ## Create facets
        facet_wrap(~survive) +
        ## Convert to circular plot
        coord_polar(theta = "x") +
        ## Annotate percentages
        geom_text(aes(y=0), size = 5) +
        theme_bw() + theme(
            panel.grid = element_blank(), panel.border = element_blank(),
            axis.text = element_blank(), axis.ticks = element_blank(), axis.title = element_blank(),
            text = element_text(size=20)
```

---

## The result!

An interactive app that is available at:

<https://dpjmullins.shinyapps.io/shiny_app/>
